Spring haul-out behavior of seals in the Bering and Chukchi seas

### 2023-10-02 analysis/paper/London_HauloutBeahvior.Rmd
more perfecting of figures/captions etc

  between 2005 and 2020 to quantify the proportion of seals hauled out on ice,

@@

  among years. However, we did not find evidence that haul-out timing was linked

@@

For this study we used haul-out behavior data and location estimates from
bio-loggers deployed on bearded, ribbon, and spotted seals in the Bering,
Chukchi, and western Beaufort seas by multiple organizations as part of
collaborative investigations from 2005 through 2020. Seals were captured using
nets and bio-loggers were attached during studies based in coastal communities
or on research ships. Ship-based capture events occurred during spring near the
southern ice edge in the Bering Sea between 2005 and 2018. Land-based capture
events occurred between 2005 and 2020 from May to October, generally between the
coastal communities of Scammon Bay, Alaska in the Bering Sea, Utqiaġvik, Alaska
in the Chukchi Sea, and Nuiqsut in the Beaufort Sea. Data from additional
deployments along the Kamchatka peninsula in the western Bering Sea are also
included. We refer readers to Figure \@ref(fig:deployplot) and the primary literature
for detailed capture and bio-logger attachment methods (see Supplemental
Material, S1).

@@

A typical model fitting exercise would also include a model selection process.
However, AIC (and similar criteria) is not suitable when using
pseudo-likelihoods, because pseudo-data generated in the model fitting process
[@verhoef2010a] differ between models [@teneyck2018]. After fitting GLMPM
models, we instead used "type III" $F$-tests to calculate $p$-values
[@verhoef2010a] to evaluate model performance and important terms. We also
produced predictions of haul-out behavior as a function of three influential
predictors (e.g. solar hour, day-of-year, age-sex). Weather covariates for these
predictions were based on daily or hourly smoothed weather covariate values
across the study region. Such predictions were then used to develop haul-out
probability surfaces, explore conditional effects of weather covariates, and
determine annual peaks in haul-out activity. The timing of peak haul-out
behavior was further used to regress against the annual maximum sea ice extent
in the study region.

Visualizing the marginal or conditional effect of an individual weather
covariate on haul-out probability was difficult in this analysis because of the
collinearity between covariates as well as the spatial and temporal variation
across such a large region. The relationship of each weather covariate with
haul-out probability, averaged over the other weather conditions, was more
variable than model coefficients would imply. That said,
important insights can be gained from plots of marginal effects. To create these
plots, we predicted haul-out probability across the full range
of weather covariate values while fixing hour of the day at local solar noon and
day-of-year at the average day of peak haul-out for each species. The
visualizations also include vertical lines representing 95% confidence intervals
around the predicted haul-out probability to better communicate the variation in
model uncertainty.

@@

```{r beardedHOCal, include=TRUE, fig.cap="Predicted hourly haul-out probability of bearded seals (all ages and sex classes) from 1 March to 15 July for all and sex classes."}

plot_df <- bearded_newdata %>%
  mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
         month = lubridate::month(date,label=TRUE),
         day = lubridate::day(date)) %>%
  filter(month != "Aug")

@@

p <- p + facet_grid(age_sex~month,
                    labeller = labeller(
                      age_sex = age_sex_labels
                    ))
p <- p + scale_x_continuous(breaks = c(1,10,20,30)) +
  scale_y_continuous(breaks = c(4,12,20), labels = c("04","12", "20"))
p <- p + theme(legend.position = "bottom",
               strip.background = element_rect(colour="white"),
               axis.ticks=element_blank()) + 
  xlab("day of month") + ylab("local solar hour") +
  labs(title = "Bearded seal predicted haul-out probability",
       subtitle = "Our model predicts lower haul-out probability for bearded seals overall 
       compared to ribbon and spotted seals. Highest probability is associated with solar noon and midnight.")
p
```

@@

Ribbon seals exhibited a pattern of gradually increasing haul-out probability
in April that peaks in late May for subadults and in early June for
adults (Figure \@ref(fig:ribbonHOCal)). The behavior was clearly centered
around local solar noon and expanded to other hours later in the spring
as seals entered their molting period. Subadults showed an earlier start
and more intense haul-out activity in April and May. The
young-of-the-year records begin after weaning and the model predictions
demonstrated the ontogeny of in-water activities (e.g. diving, foraging)
in May. Adult females had a more protracted haul-out season compared to
males, and more time was spent resting in June and July. The model suggested
adult male ribbon seals completed their molt by the end of June. Lastly,
for adults of both sexes there was some indication for a shift to a
crepuscular haul-out pattern in late June and July.

```{r ribbonHOCal, fig.asp = 0.9, include=TRUE, fig.cap="Predicted hourly haul-out probability of ribbon seals from 1 March to 15 July for each age and sex class used in the model. There is an apparent seasonal progression with subadults hauling out earlier in the season followed by adult males and, then, adult females. Predictions for young of the year show their transition from newly weaned pups resting on the ice to more in-water activities."}

plot_df <- ribbon_newdata %>%
  mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
         month = lubridate::month(date,label = TRUE),
         day = lubridate::day(date)) %>%
  filter(month != "Aug")

@@

p <- p + facet_grid(age_sex~month,
                    labeller = labeller(
                      age_sex = age_sex_labels
                    ))
p <- p + scale_x_continuous(breaks = c(1,10,20,30)) +
  scale_y_continuous(breaks = c(4,12,20), labels = c("04","12", "20"))
p <- p + theme(legend.position = "bottom",
               strip.background = element_rect(colour = "white"),
               axis.ticks = element_blank()) +
  xlab("day of month") + ylab("local solar hour") +
  ggtitle("Ribbon seal predicted haul-out probability.",
          subtitle = "Our model predicts increased haul-out probability 
          for ribbon seals centered on solar noon and a sequential 
          seasonal pattern linked to age and sex classifications.")
p
```

@@

Compared to ribbon seals, spotted seals showed a longer spring haul-out season
that was less intensely centered on solar noon (Figure \@ref(fig:spottedHOCal)).
Adults of both sexes spent considerable time in April, May and June hauled out.
In contrast to ribbon seals, adult spotted seal males had a more protracted
haul-out season compared to females, and more time out of the water in June and
July (Figures \@ref(fig:ribbonHOCal) and \@ref(fig:spottedHOCal)). As with
ribbon seals, the young-of-the-year records began after weaning and the model
predictions demonstrated the ontogeny of in-water activities (e.g. diving,
foraging) in May.

```{r spottedHOCal, fig.asp = 0.9, include=TRUE, fig.cap="Predicted hourly haul-out probability of spotted seals from 1 March to 15 July for each age and sex class used in the model. Compared to ribbon seals, the period of increased haul-out behavior is broader and there is more alignment in the seasonal timing of haul-out behavior across age and sex classifications. This likely reflects the triad behavior in spotted seals when suitor males haul out with nursing females. As with ribbon seals, redictions for young of the year show their transition from newly weaned pups resting on the ice to more in-water activities."}

plot_df <- spotted_newdata %>%
  mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
         month = lubridate::month(date,label=TRUE),
         day = lubridate::day(date)) %>%
  filter(month != "Aug")

@@

p <- p + facet_grid(age_sex~month,
                    labeller = labeller(
                      age_sex = age_sex_labels
                    ))
p <- p + scale_x_continuous(breaks = c(1,10,20,30)) +
  scale_y_continuous(breaks = c(4,12,20), labels = c("04","12", "20"))
p <- p + theme(legend.position = "bottom") +
  theme(strip.background = element_rect(colour="white")) +
  theme(axis.ticks=element_blank()) +
  xlab("day of month") + ylab("local solar hour") +
  ggtitle("Spotted seal predicted haul-out probability.",
          subtitle = "Our model predicts increased haul-out probability 
          for spotted seals centered on solar noon with general alignment in the
          seasonal timing across age and sex classifications.")
p
```

@@

```{r annualHO, fig.asp = 0.75, include=TRUE, fig.cap="Annual variability in the timing of peak haul-out probability (colored markers) for ribbon and spotted seals across 13 years. Predictions are shown for local solar noon and under smoothed weather conditions. Only those groups (age:sex + year) that included observations from more than one seal are shown. Additionally, any groups where data were only available after 1 June or before 1 May are not included."}

@@

p <-
  ggplot(plot_df %>% filter(age_sex != "YOUNG OF YEAR"), aes(yday, ho_prob)) +
  geom_pointrange(
    aes(ymin = lower95, ymax = upper95),
    alpha = 0.25,
    stroke = 0,
    color = "grey80"
  ) +
  geom_pointrange(
    data = peak_ho_df,
    mapping = aes(
      yday,
      ho_prob,
      ymin = 0,
      ymax = ho_prob,
      color = year
    ),
    shape = 15
  ) +
  scale_color_discrete(type = pal) +
  scale_x_continuous(
    breaks = c(74, 105, 135, 166),
    labels = c("15-Mar", "15-Apr", "15-May", "15-Jun")
  ) +
  facet_grid(age_sex ~ species) +
  ylab("haul-out probability") +
  labs(title = "Annual variability in the timing of peak haul-out probability",
       subtitle = "Predictions for ribbon and spotted seals are shown for local 
       solar noon and under smoothed weather conditions") +
  theme(legend.position = "bottom",
        axis.title.x = element_blank())

>>>>

  between 2005 and 2021 to quantify the proportion of seals hauled out on ice,

@@

  among years. However, we did not find broad evidence that haul-out timing was linked

@@

For this study we used haul-out behavior data and location estimates from
bio-loggers deployed on bearded, ribbon, and spotted seals in the Bering,
Chukchi, and western Beaufort seas by multiple organizations as part of
collaborative investigations from 2005 through 2021. Seals were captured using
nets and bio-loggers were attached during studies based in coastal communities
or on research ships. Ship-based capture events occurred during spring near the
southern ice edge in the Bering Sea between 2005 and 2018. Land-based capture
events occurred between 2005 and 2020 from May to October, generally between the
coastal communities of Scammon Bay, Alaska in the Bering Sea, Utqiaġvik, Alaska
in the Chukchi Sea, and Nuiqsut in the Beaufort Sea. Data from additional
deployments along the Kamchatka peninsula in the western Bering Sea are also
included. We refer readers to Figure \@ref(fig:deployplot) and the primary literature
for detailed capture and bio-logger attachment methods (see Supplemental
Material, S1).

@@

A typical model fitting exercise would also include a model selection process.
However, AIC (and similar criteria) is not suitable when using
pseudo-likelihoods, because pseudo-data generated in the model fitting process
[@verhoef2010a] differ between models [@teneyck2018]. After fitting GLMPM
models, we instead used "type III" $F$-tests to calculate $p$-values
[@verhoef2010a] to evaluate model performance and important terms. We also
produced predictions of haul-out behavior as a function of three influential
predictors (e.g. solar hour, day-of-year, age-sex). Weather covariates for these
predictions were based on daily or hourly smoothed weather covariate values
across the study region. Such predictions were then used to develop haul-out
probability surfaces, explore conditional effects of weather covariates, and
determine annual peaks in haul-out activity. The timing of peak haul-out
behavior was further used to regress against the annual maximum sea ice extent
in the study region. Predictions before 15 March and after 30 June were not
included in visualizations or other evaluations to avoid spurious model
predictions at the edge of the data range.

Visualizing the marginal or conditional effect of an individual weather
covariate on haul-out probability was difficult in this analysis because of the
collinearity between covariates as well as the spatial and temporal variation
across such a large region. The relationship of each weather covariate with
haul-out probability, averaged over the other weather conditions, was more
variable than model coefficients would imply. That said,
important insights can be gained from plots of marginal effects. To create these
plots, we predicted haul-out probability across the full range
of weather covariate values while fixing hour of the day at local solar noon and
day-of-year at 15 May. The
visualizations also include vertical lines representing 95% confidence intervals
around the predicted haul-out probability to better communicate the variation in
model uncertainty.

@@

```{r beardedHOCal, include=TRUE, fig.cap="Predicted hourly haul-out probability of bearded seals (all ages and sex classes) from 1 March to 30 June for all age and sex classes combined. Predictions in July and before 15 March were not included to avoid spurious model predictions at the edge of the data range."}

plot_df <- bearded_newdata %>%
  mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
         month = lubridate::month(date,label = TRUE),
         day = lubridate::day(date)) %>%
  filter(!month %in% c("Jul","Aug"),
         lubridate::yday(date) >73)

@@

p <- p + facet_grid(age_sex~month,
                    scales = "free_x",
                    space = "free_x",
                    labeller = labeller(
                      age_sex = age_sex_labels
                    ))
p <- p + scale_x_continuous(breaks = c(5,10,15,20,25)) +
  scale_y_continuous(breaks = c(4,12,20), labels = c("04","12", "20")) +
  coord_cartesian(expand=FALSE)
p <- p + theme(panel.spacing = unit(0.1, "lines"),
               legend.position = "bottom",
               strip.background = element_rect(colour="white"),
               axis.ticks=element_blank()) +
  xlab("day of month") + ylab("local solar hour") +
  labs(title = "Bearded seal predicted haul-out probability",
       subtitle = "Our model predicts lower haul-out probability for bearded seals overall 
       compared to ribbon and spotted seals. Highest probability is associated with solar noon and midnight.")
p
```

@@

Ribbon seals exhibited a pattern of gradually increasing haul-out probability
in April that peaks in late May for subadults and in early June for
adults (Figure \@ref(fig:ribbonHOCal)). The behavior was clearly centered
around local solar noon and expanded to other hours later in the spring
as seals entered their molting period. Subadults showed an earlier start
and more intense haul-out activity in April and May. The
young-of-the-year records begin after weaning and the model predictions
demonstrated the ontogeny of in-water activities (e.g. swimming, foraging)
in May. Adult females had a more protracted haul-out season compared to
males, and more time was spent hauled out in June compared to adult males and
subadults.

```{r ribbonHOCal, fig.asp = 0.9, include=TRUE, fig.cap="Predicted hourly haul-out probability of ribbon seals from 15 March to 30 June for each age and sex class used in the model. There is an apparent seasonal progression with subadults hauling out earlier in the season followed by adult males and, then, adult females. Predictions for young of the year show their transition from newly weaned pups resting on the ice to more in-water activities. Predictions in July and before 15 March were not included to avoid spurious model predictions at the edge of the data range."}

plot_df <- ribbon_newdata %>%
  mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
         month = lubridate::month(date,label = TRUE),
         day = lubridate::day(date)) %>%
  filter(!month %in% c("Jul","Aug"),
         lubridate::yday(date) >73)

@@

p <- p + facet_grid(age_sex~month,
                    scales = "free_x",
                    space = "free_x",
                    labeller = labeller(
                      age_sex = age_sex_labels
                    ))
p <- p + scale_x_continuous(breaks = c(5,10,15,20,25)) +
  scale_y_continuous(breaks = c(4,12,20), labels = c("04","12", "20")) +
  coord_cartesian(expand=FALSE)
p <- p + theme(panel.spacing = unit(0.1, "lines"),
               legend.position = "bottom",
               strip.background = element_rect(colour="white"),
               axis.ticks=element_blank()) +
  xlab("day of month") + ylab("local solar hour") +
  ggtitle("Ribbon seal predicted haul-out probability.",
          subtitle = "Our model predicts increased haul-out probability 
          for ribbon seals centered on solar noon and a sequential 
          seasonal pattern linked to age and sex classifications.")
p
```

@@

Compared to ribbon seals, spotted seals showed a longer spring haul-out season
that was less intensely centered on solar noon (Figure \@ref(fig:spottedHOCal)).
Adults of both sexes spent considerable time in April, May and June hauled out.
In contrast to ribbon seals, adult spotted seal males had a more protracted
haul-out season compared to females, and more time out of the water in June 
(Figures \@ref(fig:ribbonHOCal) and \@ref(fig:spottedHOCal)). As with
ribbon seals, the young-of-the-year records began after weaning and the model
predictions demonstrated the development of in-water activities (e.g. swimming,
foraging) in May.

```{r spottedHOCal, fig.asp = 0.9, include=TRUE, fig.cap="Predicted hourly haul-out probability of spotted seals from 1 March to 30 June for each age and sex class used in the model. Compared to ribbon seals, the period of increased haul-out behavior is broader and there is more alignment in the seasonal timing of haul-out behavior across age and sex classifications. This likely reflects the triad behavior in spotted seals when suitor males haul out with nursing females. As with ribbon seals, redictions for young of the year show their transition from newly weaned pups resting on the ice to more in-water activities. Predictions in July and before 15 March were not included to avoid spurious model predictions at the edge of the data range."}

plot_df <- spotted_newdata %>%
  mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
         month = lubridate::month(date,label = TRUE),
         day = lubridate::day(date)) %>%
  filter(!month %in% c("Jul","Aug"),
         lubridate::yday(date) >73)

@@

p <- p + facet_grid(age_sex~month,
                    scales = "free_x",
                    space = "free_x",
                    labeller = labeller(
                      age_sex = age_sex_labels
                    ))
p <- p + scale_x_continuous(breaks = c(5,10,15,20,25)) +
  scale_y_continuous(breaks = c(4,12,20), labels = c("04","12", "20")) +
  coord_cartesian(expand=FALSE)
p <- p + theme(panel.spacing = unit(0.1, "lines"),
               legend.position = "bottom",
               strip.background = element_rect(colour="white"),
               axis.ticks=element_blank()) +
  xlab("day of month") + ylab("local solar hour") +
  ggtitle("Spotted seal predicted haul-out probability.",
          subtitle = "Our model predicts increased haul-out probability 
          for spotted seals centered on solar noon with general alignment in the
          seasonal timing across age and sex classifications.")
p
```

@@

```{r annualHO, fig.asp = 0.75, include=TRUE, fig.cap="Annual variability in the timing of peak haul-out probability (colored markers) for ribbon and spotted seals across 14 years. The full seasonal range of annual predictions along with uncertainty are shown as grey lines. Covariates were fixed at local solar noon and under smoothed weather conditions. Only those groups (age:sex + year) that included observations from more than one seal are shown. Additionally, any groups where data were only available after 1 June or before 1 May are not included."}

@@

p <-
  ggplot(plot_df %>% filter(age_sex != "YOUNG OF YEAR"), aes(yday, ho_prob)) +
  geom_pointrange(
    aes(ymin = lower95, ymax = upper95),
    alpha = 0.25,
    stroke = 0,
    color = "grey80"
  ) +
  geom_pointrange(
    data = peak_ho_df,
    mapping = aes(
      yday,
      ho_prob,
      ymin = 0,
      ymax = ho_prob,
      color = year
    ),
    shape = 15
  ) +
  scale_color_discrete(type = pal) +
  scale_x_continuous(
    breaks = c(74, 105, 135, 166),
    labels = c("15-Mar", "15-Apr", "15-May", "15-Jun")
  ) +
  scale_y_continuous(breaks = c(0.2,0.4,0.6,0.8), labels = c("0.2","0.4", "0.6", "0.8")) +
  guides(col = guide_legend(ncol = 7)) +
  facet_grid(age_sex ~ species,
             labeller = labeller(
                      age_sex = age_sex_labels
                    )) +
  ylab("haul-out probability") +
  labs(title = "Annual variability in the timing of peak haul-out probability",
       subtitle = "Predictions of maximum annual haul-out probability for ribbon and spotted seals are indicated by the colored squares with grey lines showing the individual seasonal trajectories and uncertainty.",
       caption = "predictions are shown for local solar noon and under smoothed weather conditions") +
  theme(legend.position = "bottom",
        axis.title.x = element_blank())
